Refinement for open automata

### 2021-05-31 2021-Refinement/report.tex
Transitivity proof WIP

% transitions between sections
% TODO: CITER LES CONFERENCES

@@

\section{Composition of Open Automata}\label{sec:comp}
% TODO: distinguish composition from "parallel composition (pure interleaving)"
Open automata are partially specified automata, part of that partiality comes from the holes.
The interpretation of a hole is an interface with another open automaton, in which we can plug an open automaton with an operation called composition.
The composition of open automata was already implicitely defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never completely formalised on open automata.
The definition of composition below is a direct translation of what happens with pNets composition without the need of introducing pNets.
\begin{defi}[Composition of open automata]
The composition of \(A_c \defobject \OAg[c]\) in the hole \(k \in J_p\) of \(A_p \defobject \OAg[p]\) is
\begin{align*}
	A_p\subst{A_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{with } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p} \in T_p, k \notin J'_p, s_c \in S_c}
\end{align*}
\end{defi}
The action emitted when \(A_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(A_p\) which have it as a hole action (first transition set, \(\alpha_c = \beta_k\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) however it is asymmetric because of the second transition set (\(k \notin J'_p\)):
No transition of \(A_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(A_p\).

@@

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}

The important properties we consider for the relation on automata are:
\begin{defi} A relation on open automata \(\leq\) is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a\subst{b}{j} \leq a\);
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \cong b\subst{c}{j}\);
\defitem \textbf{context refining for composition} iff \(\forall a\, b\, c, a \leq b \implies a\subst{c}{j} \leq b\subst{c}{j}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\subst{a}{j} \leq c\subst{b}{j}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\subst{a}{j} \leq d\subst{b}{j}\);
\defitem \textbf{compatible with FH-bisimulation} iff \(\forall a\, b\, c, d, a \cong b \wedge c \cong d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Correctness w.r.t.\@ composition means that every composition is considered a refinement.
This property captures the expected behaviour of a refinement relation on a compositionnal structure like the open automata.
Completeness w.r.t.\@ composition means that a refinement corresponds to the left automaton being equivalent to some composition of the right automaton.
This property is useful if we want to characterise the kind of operations that can generate any refined automaton starting from the specification.
We won't have it because extending the domain of the initial valuation will be a refinement which can't be expressed with composition.
Yet a ``completeness w.r.t.\@ some set of operations" is still a strong and interesting property.
Context refinement is the refinement being compatible with composition of the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the conjunction of the two latter (assuming the relation is a preorder).
This property is required on compositionnal structures to be of any use in verification, proofs and for model checking.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.
Since FH-bisimulation is the pre-existing equivalence relation, we should not introduce an incompatible proof strategy.

The important properties we consider for the relation on states are being a simulation and the reduction of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the holes actions from same hole names (\(J'_1 \cap J'_2\)) match (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables and holes and guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i, 1 \leq i \leq n\) in \(A_0\) gives the holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{k_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

Deadlocks are configurations reachable in a run and from which no further transition is possible.
They can arise unpredictably when parallelism is involved (which is the case in our composition) and they break the intended behaviour of a system.
By preventing them we could hope that the refinement preserves some liveness properties.
\begin{defi}[Deadlock reduction, intuitive definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if:
\defitem It is a pre-simulation;
\defitem If from related states, there is a possible transition in the specification, then there is a pair of matching transition:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \to \values, \mpar{\sigma \vdash R\mpar{s_1, s_2}} \implies \\
	\mpar{\exists t_2 \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \vdash \fguard{t_2}} \implies \\
	\bigsymb{\exists} \mpar{t_1, t_2} \defobject \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2}, \exists \nu: \fvars{t_1} \uplus \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This requirement effectively prevents the presence of deadlock state in the first automaton if there is a related non-deadlock state in the second by contraposition.
On top of that the transition preventing the deadlock has to be an existing transition in the specification.
This is because a valid transition unmatched in the specification is not a deadlock, yet we do not want it.
The direct formulation of the property would also have required the unintuitive quantification on a possibly empty set of transition to state that they are impossible.
The two previous reasons motivated the formulation in contraposed form.

``Deadlock reduction" conflicts with all the properties where composition is involved (for instance ``composition correctness") by disallowing some unwanted cases where an automaton in a hole cannot produce any action that any out-transition expects.
In particular filling a hole with the deadlock automaton (only one state without out-transitions) is not considered a refinement for a deadlock reducing relation.
A way to solve these conflicts is to characterise a composition that do not introduce deadlocks, which will be used instead of the composition introduced earlier.
\begin{defi}[Reachability]
For any open automata \(A \defobject \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is a predicate on states such that
\defitem \(\sigma_0 \vdash \reach{A}\mpar{s_0}\)
\defitem \nmm{\bigsymb{\forall} t \defobject \OTg \in T, \forall \nu: \fvars{t} \to \values, \nu \vdash \reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}}
\end{defi}
For the reader not used to simulations and bisimulations, this definition of a predicate can seem strange.
The reachability predicate is used to characterise the states and valuations that are reachable in a run of an automaton.
In fact the role of the predicate is to characterise potentially reachable configurations without having to characterise configurations as the result of a valid path in an automaton.

To do that we impose that the initial configuration is reachable and that reachability is preserved by taking valid transitions.
This effectively makes reachability account for all paths and potentially over-approximate the reachable configurations.
However the exact reachability may not be reprensentable in the formulas, hence the need of potentially over-approximating.

Another way of understanding this predicate is that it is a fixpoint of the union of valuations (function \(f\) defined below) that contains the initial valuation:
\begin{align*}
	f\mpar{p} = \mbrc{\mbox{\nmm{s' \mapsto p\mpar{s'} \vee \bigsymb{\bigvee_\subbox{\OTg \,\in\, \fIT{s'}}} p\mpar{s}\psubst{\psi} \wedge g}}} && % TODO: fix that in some way
	\sigma_0 \vdash \reach{A}\mpar{s_0} &&
	f\mpar{\reach{A}} = \reach{A}
\end{align*}

\begin{defi}[Non-locking composition, intuitive definition]
Let \(A_p\subst{A_c}{j} = \OAg\) with \(A_c \defobject \OAg[c]\), \(A_p \defobject \OAg[p]\) and \(j \in J_p\).
The composition is non-locking if there is a reachability predicate such that from reachable configurations, if there was a possible transition in \(A_p\) then there is a possible transition in \(A_p\subst{A_c}{j}\):
\begin{multline*}
	\forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{j}}\mpar{s}} \implies \\
	\mpar{\exists t_p \in \fOT{s_p}, \exists \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \vdash \fguard{t_p}} \implies \\
	\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}
\end{multline*}
\end{defi}
An illustration of the introduction of a deadlock by composition and the definition of non-locking composition is given in Appendix \ref{apx:lockcomp}.

We can expand the aliases \(S, V\) into \(S_p \times S_c, V_p \uplus V_c\) and the last transition \(t\) into a pair \(\mpar{t_p, t}\) where the first transition is the one from \(A_p\) which generated \(t\).
Doing that makes the definition become mildly similar with the deadlock reduction one.
It is not a coincidence because their goal is to caracterise the same kind of compatibility between automata.
Actually it is possible to simplify both definition and make them even more similar.
\begin{defi}[Deadlock reduction, working definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if it is a pre-simulation and:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \to \values, \\
	\sigma \vdash \mpar{R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{defi}
The requirement here is that from any related configurations, if there is a possible transition in the second automaton (whatever the value of the free variables) then there is a possible transition in the first automaton (the existential quantifier is implicit in the notation \(\vdash\)).
The transition are not matched because the relation on configurations is a pre-simulation and it is sufficient to ensure it as will be shown in the proof.
Apart from the matching transitions this is exactly the same explaination as the intuitive definition.
The expected advantage of this formulation is that a SMT should behave better without the quantifiers.
Indeed the big disjunctions are not quantifiers because in practise the out-transitions are finitary, and so is the expansion of the disjunction.
\begin{lem}
The intuitive and working definition of deadlock reduction are equivalent.
\end{lem}

\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{j}\) is a non-locking composition if:
\[ \forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \uplus \biguplus_\subbox{t_p \in \fOT{s_p}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{s_p}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
Proof of both lemma are given in Appendix \ref{apx:lemeqd}.

This new definition of composition can replace the standard composition in the case where only one hole is filled, and also in the parallel case.
In the parallel case, we may think that if the condition is checked in the parallel case (at the final composition), then the composition is non-locking for every composition order.
It is not true.
Actually even if the last composition of any order of composition is non-locking there is no guarantee that this order is non-locking at every step.
However the oppsite way is true even if uninformative:
If there is an order of composition where every composition is non-locking then the last composition is non-locking, last composition resulting in the same automaton as the parallel composition.
So the condition is better checked after the parallel composition or the last individual composition.
\begin{exi}
\begin{figure}
\centering
\input{file_sync.tex}
\caption{A made-up protocol for file synchronisation}
\label{fig:pnls}
\end{figure}
The automaton on Figure \ref{fig:pnls} is a sample protocol for file synchronisation.
The two holes are \(net\) a network manager used to perform operations on the network, and \(hash\) a file integrity check used to generate a fingerprint.
The state \(i\) is the initial state.
In that state we want to synchronise the files before working, or work on the local version as a fallback.
If the preferred version is the remote one, we go in state \(d1\) asking the netwotk manager to fetch the lastest version, then ask the integrity checker to validate the fingerprint.
If the preferred version is the local one, we go in state \(d2\) asking for the integrity checker to generate a fingerprint for the local files, then ask the network manager to push them.
In fallback we use the local version.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref} % TODO: Fold definitions and proofs
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\psubst{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{i \in \mdbrk{1; n}}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

>>>>

% transitions between sections
% TODO: CITER LES CONFERENCES
% change deadlock reduction into deadlock prevention?

@@

\section{Composition of Open Automata}\label{sec:comp}
Open automata are partially specified automata, part of that partiality comes from the holes.
The interpretation of a hole is an interface with another open automaton, in which we can plug an open automaton with an operation called composition.
In this article, composition will only refer to filling holes, other compositions like ``parallel composition" (pure interleaving) are not tackled.
The composition of open automata was already implicitely defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never completely formalised on open automata.
The definition of composition below is a direct translation of what happens with pNets composition without the need of introducing pNets.
\begin{defi}[Composition of open automata]
The composition of \(A_c \defobject \OAg[c]\) in the hole \(k \in J_p\) of \(A_p \defobject \OAg[p]\) is
\begin{align*}
	A_p\subst{A_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{with } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p} \in T_p, k \notin J'_p, s_c \in S_c}
\end{align*}
\end{defi}
The action emitted when \(A_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(A_p\) which have it as a hole action (first transition set, \(\alpha_c = \beta_k\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) however it is asymmetric because of the second transition set (\(k \notin J'_p\)):
No transition of \(A_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(A_p\).

@@

Two states \(s_1 \in S_1, s_2 \in S_2\) with their respective valuations \(\sigma_1: V_1 \to \values, \sigma_2: V_2 \to \values\) are related iff \(\sigma_1 \uplus \sigma_2 \vdash R\mpar{s_1, s_2}\).
\end{defi}
The important properties we consider for the relation on automata are:
\begin{defi} A relation on open automata \(\leq\) is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a\subst{b}{j} \leq a\);
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \cong b\subst{c}{j}\);
\defitem \textbf{context refining for composition} iff \(\forall a\, b\, c, a \leq b \implies a\subst{c}{j} \leq b\subst{c}{j}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\subst{a}{j} \leq c\subst{b}{j}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\subst{a}{j} \leq d\subst{b}{j}\);
\defitem \textbf{compatible with FH-bisimulation} iff \(\forall a\, b\, c, d, a \cong b \wedge c \cong d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Correctness w.r.t.\@ composition means that every composition is considered a refinement.
This property captures the expected behaviour of a refinement relation on a compositionnal structure like the open automata.
Completeness w.r.t.\@ composition means that a refinement corresponds to the left automaton being equivalent to some composition of the right automaton.
This property is useful if we want to characterise the kind of operations that can generate any refined automaton starting from the specification.
We won't have it because extending the domain of the initial valuation will be a refinement which can't be expressed with composition.
Yet a ``completeness w.r.t.\@ some set of operations" is still a strong and interesting property.
Context refinement is the refinement being compatible with composition of the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the conjunction of the two latter (assuming the relation is a preorder).
This property is required on compositionnal structures to be of any use in verification, proofs and for model checking.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.
Since FH-bisimulation is the pre-existing equivalence relation, we should not introduce an incompatible proof strategy.

The important properties we consider for the relation on states are being a simulation and the reduction of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

Deadlocks are reachable configurations and from which no further transition is possible.
They can arise unpredictably when parallelism is involved (which is the case in our composition) and they break the intended behaviour of a system.
By preventing them we could hope that the refinement preserves some liveness properties.
\begin{defi}[Deadlock reduction, intuitive definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if:
\defitem It is a pre-simulation;
\defitem If from related states, there is a possible transition in the specification, then there is a pair of matching transition:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \to \values, \mpar{\sigma \vdash R\mpar{s_1, s_2}} \implies \\
	\mpar{\exists t_2 \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \vdash \fguard{t_2}} \implies \\
	\bigsymb{\exists} \mpar{t_1, t_2} \defobject \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2}, \exists \nu: \fvars{t_1} \uplus \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
Actually, this requirement prevents the presence of deadlock state in the first automaton if there is a related non-deadlock state in the second by contraposition.
On top of that the transition preventing the deadlock has to be an existing transition in the specification.
This is because a valid transition unmatched in the specification is not a deadlock, yet we do not want it.
The direct formulation of the property would also have required the unintuitive quantification on a possibly empty set of transition to state that they are impossible.
The two previous reasons motivated the formulation in contraposed form.

``Deadlock reduction" conflicts with all the properties where composition is involved (for instance ``composition correctness") by disallowing some unwanted cases where an automaton in a hole cannot produce any action that any out-transition expects.
In particular filling a hole with the deadlock automaton (only one state without out-transitions) is not considered a refinement for a deadlock reducing relation.
A way to solve these conflicts is to characterise a composition that do not introduce deadlocks, which will be used instead of the composition introduced earlier.
\begin{defi}[Reachability]
For any open automata \(A \defobject \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is a predicate on states satisfying both
\defitem Inital state is reachable: \(\sigma_0 \vdash \reach{A}\mpar{s_0}\)
\defitem Reachability is preserved across transitions: \nmm{\bigsymb{\forall} t \defobject \OTg \in T, \forall \nu: \fvars{t} \to \values, \nu \vdash \reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}}
\end{defi}
The reachability predicate is used to characterise the reachable configurations in a run of an automaton.
In fact the role of the predicate is to characterise potentially reachable configurations without having to characterise configurations as the result of a valid path in an automaton.

To do that we impose that the initial configuration is reachable and that reachability is preserved by taking valid transitions.
This effectively makes reachability account for all paths and potentially over-approximate the reachable configurations.
The exact reachability may not be reprensentable in the formulas, hence the need of potentially over-approximating. % TODO: is representable the correct word

% Another way of understanding this predicate is that it is a fixpoint of the union of valuations (function \(f\) defined below) that contains the initial valuation:
% \begin{align*}
	% f\mpar{p} = \mbrc{\mbox{\nmm{s' \mapsto p\mpar{s'} \vee \bigsymb{\bigvee_\subbox{\OTg \,\in\, \fIT{s'}}} p\mpar{s}\psubst{\psi} \wedge g}}} && % TODO: fix that in some way
	% \sigma_0 \vdash \reach{A}\mpar{s_0} &&
	% f\mpar{\reach{A}} = \reach{A}
% \end{align*}
\begin{defi}[Non-locking composition, intuitive definition]
Let \(A_c \defobject \OAg[c]\), \(A_p \defobject \OAg[p]\) and \(j \in J_p\).
The composition \(A \defobject A_p\subst{A_c}{j}\) is non-locking if \(A\) has a reachability predicate satisfying:
From reachable configurations, if there was a possible transition in \(A_p\) then there is a possible transition in \(A_p\subst{A_c}{j}\).

In formula, for \(A = \OAg\), it gives
\begin{multline*}
	\forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{j}}\mpar{s}} \implies \\
	\mpar{\exists t_p \in \fOT{s_p}, \exists \nu_p: \fvars{t_p} \to \values, \sigma \uplus \nu_p \vdash \fguard{t_p}} \implies \\
	\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}
\end{multline*}
\end{defi}
We can expand the aliases \(S, V\) into \(S_p \times S_c, V_p \uplus V_c\) and transform the last transition \(t\) into a pair \(\mpar{t_p, t}\) where the first transition is the one from \(A_p\) which generated \(t\).
Doing that makes the definition become mildly similar with the deadlock reduction one. % OR slightly
It is not a coincidence because their goal is to caracterise the same kind of compatibility between automata.
Actually it is possible to simplify both definition and make them even more similar.
\begin{defi}[Deadlock reduction, working definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if it is a pre-simulation and it satisfies the following:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \to \values, \\
	\sigma \vdash \mpar{R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{defi}
The requirement here is that from any related configurations, if there is a possible transition in the second automaton (whatever the value of the free variables) then there is a possible transition in the first automaton (the existential quantifier is implicit in the notation \(\vdash\)).
The transition are not matched because the relation on configurations is a pre-simulation and it is sufficient to ensure it as will be shown in the proof.
Apart from the matching transitions this is exactly the same explaination as the intuitive definition.
The expected advantage of this formulation is that a SMT should behave better without the quantifiers.
Indeed the big disjunctions are not quantifiers because in practise the out-transitions are finitary, and so is the expansion of the disjunction. % WHAT IS THE ISSUE HERE?
\begin{lem}
The intuitive and working definition of deadlock reduction are equivalent.
\end{lem}
\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{j}\) is a non-locking composition if:
\[ \forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \uplus \biguplus_\subbox{t_p \in \fOT{s_p}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{s_p}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
An illustration of the introduction of a deadlock by composition and the definition of non-locking composition is given in Appendix \ref{apx:lockcomp}.
Proof of both lemma are given in Appendix \ref{apx:lemeqd}.

This new definition of composition can replace the standard composition in the case where only one hole is filled, and also be extended to a parallel composition.
For the parallel composition, we may think that if the composition is non-locking, then every composition is non-locking for all composition order.
It is not true.
Actually even if the last composition of any order of composition is non-locking there is no guarantee that this order is non-locking at each step.
However the oppsite way is true even if uninformative: % WHAT IS THE ISSUE?
If there is an order of composition where every composition is non-locking then the last composition is non-locking, last composition resulting in the same automaton as the parallel composition.
So the condition is better checked after the parallel composition or the last individual composition.
\begin{exi}
\begin{figure}
\centering
\input{file_sync.tex}
\caption{A made-up protocol for file synchronisation}
\label{fig:pnls}
\end{figure}
The automaton on Figure \ref{fig:pnls} is a sample protocol for file synchronisation.
The two holes are \(net\) a network manager used to perform operations on the network, and \(hash\) a file integrity check used to generate a fingerprint.
The state \(i\) is the initial state.
In that state we want to synchronise the files before working, or work on the local version as a fallback.
If the preferred version is the remote one, we go in state \(d1\) asking the netwotk manager to fetch the lastest version, then ask the integrity checker to validate the fingerprint.
If the preferred version is the local one, we go in state \(d2\) asking for the integrity checker to generate a fingerprint for the local files, then ask the network manager to push them.
In fallback we use the local version.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\psubst{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{i \in \mdbrk{1; n}}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.
